---
title: "BIO 445 Challenge Week 1"
author: "Kyle Nessen"
date: "2023-09-26"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Step 3 Question

**Prompt:** Create 3 vectors. Give the vectors logical names of your
choosing - Vector one contains the integers 10 through 15 - Vector two
contains odd numbers 3 through 13. You can manually type out these
numbers and combine them into a vector. For an extra challenge try to
use the seq() function. The arguments inside the seq function that you
will need to specify are from, to, and by. See this website
<https://www.statology.org/seq-function-in-r/> for examples and help.
You can also find details for seq() on the cheatsheet handed out in
class. - Vector three contains the product (multiplication) of vector 1
and vector 2

## Vector one

```{r}
vector_one <- 10:15 # creates a variable with the integers 10 through 15
vector_one # prints vector_one
```

## Vector two

```{r}
vector_two <- seq(from=3, to=13, by=2) #starts at 3, and skips even numbers till it reaches 13
vector_two

```

## Vector three

```{r}
vector_three <- vector_one * vector_two # multiply the two variables
vector_three
```

# Step 4 Question

**Prompt:** Join the 3 vectors together to create a two-dimensional data
structure where the first vector is column 1, the second vector is
column 2, and the third vector is column 3

```{r}
df <- data.frame("column1" = vector_one, "column2" = vector_two, "column3" = vector_three) # create a dataframe
df

```

# Step 5 Question

**Prompt:** Write a line of code to determine what type of object you
have. Is it a matrix or data frame?

```{r}
class(df) # check the class of my df variable
```

# Step 6 Question

**Prompt:** Write a line of code to extract the dimensions of your
two-dimensional object. Hint: like many things in R, there are multiple
ways to do this. Can you get this information with a single function?
What is another way to get the same information using two separate
functions?

```{r}
dim(df) # returns the dimensions of the dataframe (rows, columns)
```

# Step 7 Question

**Prompt:** Write code to: - Find the mean of column 3 - Find the
maximum value of row 1

## Mean of column 3

```{r}
mean(df$column3) # calculate mean of column 3
```

## Maximum value of row 1

```{r}
row_1_values <- df[1,]
max(row_1_values)
```

# Step 8 Question

**Prompt:** Install the palmerpenguins package. We will use this package
in next week's challenge. Load the newly installed package using the
library() function. Write a line of code to use the Help to learn about
palmerpenguins.

## Install palmerpenguins

```{r}
install.packages('palmerpenguins')
```

## load library and run Help function

```{r}
library(palmerpenguins)
help(package='palmerpenguins')
```



