---
title: "R Mini Challenge 2"
author: "Kyle Nessen"
date: "2023-10-03"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Step 1

*Load the palmerpenguins package using the library() function. If you
did not complete last week's challenge, you will need to install this
package. Load the associated dataset using data(penguins)*

```{r}
library(palmerpenguins) # load palmerpenguins library
```

# Step 2

*Familiarize yourself with the penguins data frame. - Write a line of
code to examine the first 6 rows of penguins. - How many columns are
there? You can see this when looking at the output of the previous step,
but you can also write code to extract this information. What function
would you use? - Examine the structure of the data frame. You will need
to figure out which function to use to do this. How many columns contain
integers? - When looking at the structure of penguins, you should be
able to see that multiple variables in the data frame are stored as
factors. We have not discussed these yet. Factors are really important
when performing statistical analyses in R as they represent categorical
variables with different levels. Sometimes we need to convert data
stored as characters into factors before using the data in a statistical
test. Read about factors in R here
(<https://swcarpentry.github.io/r-novice-inflammation/12-supp-factors/>)*

Loading the penguin data and looking ate first few rows

```{r}
df <- penguins # load penguins data as "df"
head(df,6) #return first 6 rows
```

Counting number of columns in the dataframe

```{r}
ncol(df) # Returns the number of columns (8)
```

Returning all columns that are integers

```{r}
# Identify columns that are integers
intCols <- sapply(df, is.integer)

# Pull names of columns that are integers
integer_column_names <- names(intCols)[intCols]

# Print names
print(integer_column_names)

```

Returning all columns that are factors.

```{r}
# Identify columns that are factors
factorCols <- sapply(df, is.factor)

# Extract names of columns where is.factor is TRUE
factor_column_names <- names(factorCols)[factorCols]

# Print the names of factor columns
print(factor_column_names)

```

Factors are a datatype that captures categorical information, such as
sex, or species name. They are different than normal string data which
can have any textual information.

# Step 3

*It is not uncommon for large data sets to contain missing values or
NAs. It is important to be aware of these missing values before
proceeding with any analysis. Use the summary() function to identify
which columns in penguins contains NAs. Write an annotation in your
script to make note of what you find.*

Running summary() on the data

```{r}
summary(df)
```

It appears that `bill_length_mm`, `bill_depth_mm`, `flipper_length_mm`,
and `body_mass_g` all have two missing values (NA), while `sex` has 11
missing values.

# Step 4

*Recall that one way of extracting a column from a data frame is to use
the following notation dataframe\$column. For example, we can extract
the column containing penguin bill length measurements as follows:
penguins\$bill_length_mm Write three lines of code to extract the
columns for the 3 other morphometric variables (body mass, flipper
length, bill depth)? Be careful and pay close attention to the way the
column names are formatted. Remember that R needs them to match
perfectly!*

```{r}
df$body_mass_g
df$body_mass_g
df$flipper_length_mm
```

# Step 5

*We often want to examine the distribution of a continuous variable
before testing for any statistical relationships. R contains a large
number of ways to plot and visualize data. In base R (no packages
needed), the hist() function will make a basic histogram, which is great
for exploring data (less great if making a figure for a final report or
publication).*

-   *Use the following code to make a histogram of penguin bill length:*

    -   *`hist(penguins$bill_length_mm)`*

-   *Now, write code to make 3 additional histograms for the remaining 3
    morphometric variables (body mass, flipper length, and bill depth).*

-   *How would you describe the shape of the histogram for each? Are the
    variables normally distributed?*

## Bill Length

Perhaps slightly bimodal?

```{r}
hist(df$bill_length_mm) # Plot data distribution
shapiro.test(df$bill_length_mm) # Testing for normalcy (not normal)
```

## Body Mass

Skewed right

```{r}
hist(df$body_mass_g) # Plot data distribution
shapiro.test(df$body_mass_g) # Testing for normalcy (not normal)
```

## Bill Depth

Bimodal with a skew to the left

```{r}
hist(df$bill_depth_mm) # Plot data distribution
shapiro.test(df$bill_depth_mm) # Testing for normalcy (not normal)
```

# Step 6

*Another useful type of plot for exploring data is a scatterplot which
can be used to visualize the relationship between two continuous
variables. For example, we might expect that penguins with longer bill
also have deeper bills.*

*• Use the following code to visualize this relationship:*

*`plot(y = penguins$bill_length_mm, x = penguins$bill_depth_mm)`*

```{r}
plot(y = df$bill_length_mm, x = df$bill_depth_mm)
```

*• Notice that we have specified which variable we want on the x-axis
and which we want on the y-axis inside the plot() function.*

*• What do you think after examining this plot? Is there visual evidence
of a positive correlation between the two variables?*

From first glance, I do not see a correlation.

*• We may also expect that the body mass and flipper length of penguins
are positively correlated. Make a scatterplot to visualize this
relationship. Put body mass on the x-axis and flipper length on the
y-axis.*

```{r}
plot(x = df$body_mass_g, y = df$ flipper_length_mm)
```

*• What conclusion do you reach after examining this plot? Does there
appear to be a positive correlation between these two variables? Note:
this relationship should be a lot easier to interpret than the first
scatterplot of bill morphometrics.*

Yes, this graph seems to show a clear positive correlation, where
flipper size increases with body mass

# Step 7

*Sometimes we want to explore the relationship between a continuous
variable and a categorical variable. For example, we may be curious as
to whether penguins are sexually dimorphic. Another way of phrasing this
is: do male and female penguins have different sized bodies? We can use
a boxplot to explore this relationship.*

*• Try the following line of code:*

*`boxplot(penguins$body_mass_g ~ penguins$sex)`*

```{r}
boxplot(df$body_mass_g ~ df$sex)
```

*• The \~ symbol in R can be read "as a function of", so this boxplot is
showing body mass as a function of sex. Note that the continuous
variable is specified first followed by the categorial variable
(factor).*

*• The thick black lines on the boxplot show the median value. Using the
plot, estimate the median body mass of males and females. Write an
annotation in your script that lists your estimates.*

**Female median value:** 3550

**Male median value:** 4200

*• Check your estimates using code. There are many ways to do this. One
way that we have seen before (refer to Intro to R script from week 1,
see line 269 and on) is to use the subset() function to extract all the
data for female penguins and then a second line of code to find the
median body mass. See if you can write out in words what each line is
doing:*

*`females <- subset(penguins, sex == "female" )`*

*`median(females$body_mass_g)`*

```{r}
females <- subset(df, sex == "female" ) # create a new dataframe that only has female data
median(females$body_mass_g) # calculates median for body mass
```

*Use the copy, paste, edit approach to find the median body mass for
male penguins*

```{r}
males <- subset(df, sex == "male" ) # create a new dataframe that only has male data
median(males$body_mass_g) # calculates median for body mass
```

# Step 8

*This data set contains a number of other categorical variables. Next,
we will explore the relationship between body morphometrics and species.
There are three penguin species in Antarctica: Adelie, Chinstrap, and
Gentoo penguins. Let's figure out which species is largest.*

*• Make a boxplot of flipper length as a function of species*

```{r}
boxplot(df$flipper_length_mm~df$species)
```

*• Make a 2nd boxplot of body mass as a function of species*

```{r}
boxplot(df$body_mass_g~df$species)
```

*• Which species is largest based on these two morphometrics?*

**Gentoo**

*• Write a line of code to subset the data so that it only contains the
species you have identified as the largest.*

```{r}
gentoo = subset(df,df$species == "Gentoo")
head(gentoo)
```

*• Examine the summary for this reduced data frame that only contains
the largest species.*

```{r}
summary(gentoo)
```

*Notice that the data contains NAs!*

*• Write a line of code to find the mean body mass for this species. You
will need to add the argument na.rm = T inside the function to ignore
the NA. Use the help in the lower right pane of RStudio for guidance.*

```{r}
mean(gentoo$body_mass_g, na.rm = TRUE)
```

*• Look at the boxplot for body mass by species. If we performed a
statistical test to compare the average body mass of each species (e.g.,
an ANOVA), what do you think the conclusions would be? Write three
sentences that state the relationship for each pairwise combination of
penguin species (1. Adelie/Gentoo, 2. Adelie/Chinstrap, 3.
Chinstrap/Gentoo). Relationships can be greater than, less than, or
equal. Here is an example based on flipper length for Adelie compared
with Gentoo: "The average flipper of Adelie penguins is less than the
average flipper length of Gentoo penguins"*

```{r}
boxplot(df$body_mass_g~df$species)
```

1.  The average body mass of Gentoo is greater than Chinstrap
2.  The average body mass of Gentoo is greater than Adelie
3.  The average body mass of Chinstrap is equal to Adelie
